---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapmixture

<!-- badges: start -->
[![R-CMD-check](https://github.com/Tom-Jenkins/mapmixture/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Tom-Jenkins/mapmixture/actions/workflows/R-CMD-check.yaml)
<!-- [![Codecov test coverage](https://codecov.io/gh/Tom-Jenkins/mapmixture/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Tom-Jenkins/mapmixture?branch=main) -->
<!-- badges: end -->

`mapmixture` is an R package and shiny app that enables users to visualise admixture as pie charts on a projected map. It also allows users to visualise admixture as traditional structure barplots or facet barplots.

## Installation

You need to have [R](https://www.r-project.org/) (>= 4.1.0) installed on your system. Click [here](https://cran.r-project.org/bin/windows/base/) to download the latest version of R for Windows.

Install the latest development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("Tom-Jenkins/mapmixture")
```

## Examples

Standard usage of `mapmixture()`:

```{r}
# Load package
library(mapmixture)

# Read in admixture file format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

# Read in coordinates file
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

# Run mapmixture
map1 <- mapmixture(admixture1, coordinates, crs = 3035)
map1
```

Customised usage of `mapmixture()`:

```{r}
# Load package
library(mapmixture)

# Read in admixture file format 3
file <- system.file("extdata", "admixture3.csv", package = "mapmixture")
admixture3 <- read.csv(file)

# Read in coordinates file
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

# Run mapmixture
map2 <- mapmixture(admixture3, coordinates,
  cluster_cols = c("blue","green"),
  cluster_names = c("Group A","Group B"),
  crs = 3035,
  boundary = c(xmin=-15, xmax=16, ymin=40, ymax=62),
  pie_size = 1,
  pie_border = 0.3,
  pie_opacity = 1,
  land_colour = "#d9d9d9",
  sea_colour = "#deebf7",
  expand = TRUE,
  arrow = TRUE,
  arrow_size = 1.5,
  arrow_position = "tl",
  scalebar = TRUE,
  scalebar_size = 1.5,
  scalebar_position = "tl",
  plot_title = "Mapmixture Figure",
  plot_title_size = 12,
  axis_title_size = 10,
  axis_text_size = 8
)
map2
```

Add additional geoms or theme options to `mapmixture()` ggplot object:
```{r}
# Load packages
library(mapmixture)
library(ggplot2)

# Read in admixture file format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

# Read in coordinates file
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

# Run mapmixture
map3 <- mapmixture(admixture1, coordinates,
  cluster_cols = c("#91bfdb","#fc8d59"),
  cluster_names = c("Ancestry 1","Ancestry 2"),
  crs = 4326,
  boundary = c(xmin=-15, xmax=16, ymin=40, ymax=62),
  pie_size = 2,
)+
  # Add additional label to the map
  annotate("label",
    x = -10,
    y = 46.5,
    label = "Atlantic Ocean",
    size = 3,
  )+
  # Add additional text to the map
  annotate("text",
    x = 2.5,
    y = 57,
    label = "North Sea",
    size = 3,
  )+
  # Adjust ggplot theme options
  theme(
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
  )+
  # Adjust the size of the legend keys
  guides(fill = guide_legend(override.aes = list(size = 5, alpha = 1)))
map3
```

Combine admixture mapmixture and barplot ggplot objects into a single figure:
```{r}
# Load packages
library(mapmixture)
library(ggplot2)
library(gridExtra)

# Read in admixture file format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

# Read in coordinates file
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

# Run mapmixture
map4 <- mapmixture(admixture1, coordinates,
  cluster_cols = c("#91bfdb","#fc8d59"),
  cluster_names = c("Ancestry 1","Ancestry 2"),
  crs = 4326,
  boundary = c(xmin=-20, xmax=20, ymin=40, ymax=62),
  pie_size = 2.5,
)+
  # Adjust theme options
  theme(
    legend.position = "top",
  )+
  # Adjust the size of the legend keys
  guides(fill = guide_legend(override.aes = list(size = 5, alpha = 1)))

# Traditional structure barplot
structure_barplot <- structure_plot(admixture1,
  type = "structure",
  cluster_cols = c("#91bfdb","#fc8d59"),
  site_dividers = TRUE,
  labels = "site",
  flip_axis = FALSE,
  site_ticks_size = -0.05,
  site_labels_y = -0.35,
  site_labels_size = 2.5
)+
  # Adjust theme options
  theme(
    axis.title.y = element_text(size = 8, hjust = 1),
    axis.text.y = element_text(size = 5),
  )

# Arrange plots
grid.arrange(map4, structure_barplot, nrow = 2, heights = c(4,1))

# Facet structure barplot
facet_barplot <- structure_plot(admixture1,
  type = "facet",
  cluster_cols = c("#91bfdb","#fc8d59"),
  facet_col = 2,
  ylabel = "Admixture proportions",
)+
  theme(
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 5),
    strip.text = element_text(size = 6, vjust = 1, margin = margin(t=1.5, r=0, b=1.5, l=0)),
  )

# Arrange plots
grid.arrange(map4, facet_barplot, ncol = 2, widths = c(3,2))
```

Use a raster as the basemap:  

The raster (TIFF) used in the example below was downloaded from Natural Earth [here](https://www.naturalearthdata.com/downloads/50m-raster-data/50m-natural-earth-1/). You need to install the [terra](https://github.com/rspatial/terra) package to use this feature. Currently, the `basemap` argument only accepts a `SpatRaster` object.

```{r}
# Load packages
library(mapmixture)
library(terra)

# Create SpatRaster object
earth <- terra::rast("../NE1_50M_SR_W/NE1_50M_SR_W.tif")

# Read in admixture file format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

# Read in coordinates file
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

# Run mapmixture
map5 <- mapmixture(admixture1, coordinates, crs = 3035, basemap = earth)
map5
```

## Launch interactive Shiny app

```{r eval=FALSE}
# Load package
library(mapmixture)

# Launch shiny app
launch_mapmixture()
```

## Link to online Shiny app
https://tomjenkins.shinyapps.io/mapmixture/

## Format

```{r}
# Load package
library(mapmixture)

# Admixture Format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)
head(admixture1)

# Admixture Format 2
file <- system.file("extdata", "admixture2.csv", package = "mapmixture")
admixture2 <- read.csv(file)
admixture2

# Admixture Format 3
file <- system.file("extdata", "admixture3.csv", package = "mapmixture")
admixture3 <- read.csv(file)
admixture3

# Coordinates
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)
coordinates
```

## Related software

- [pophelper](https://github.com/royfrancis/pophelper)
- [pophelperShiny](https://github.com/royfrancis/pophelperShiny)

