---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapmixture

<!-- badges: start -->
[![R-CMD-check](https://github.com/Tom-Jenkins/mapmixture/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Tom-Jenkins/mapmixture/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`mapmixture` is an R package and shiny app that enables users to visualise admixture as pie charts on a projected map or as traditional STRUCTURE barplots or facet barplots.

## Installation

You need to have [R](https://www.r-project.org/) (>= 4.1.0) installed on your system. Click [here](https://cran.r-project.org/bin/windows/base/) to download the latest version of R for Windows.

Install the latest development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("Tom-Jenkins/mapmixture", dependencies = TRUE)
```

## Examples

Run `mapmixture()` with using admixture format 1 and custom parameters:

```{r}
# Load package
library(mapmixture)

# Admixture Format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

# Coordinates
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

mapmixture(admixture1, coordinates,
  cluster_cols = c("blue","green"),
  cluster_names = c("Group 1","Group 2"),
  crs = 3035,
  boundary = c(xmin=-15, xmax=16, ymin=40, ymax=62),
  pie_size = 1,
  pie_border = 0.3,
  pie_opacity = 1,
  land_colour = "#d9d9d9",
  sea_colour = "#deebf7",
  expand = TRUE,
  arrow = TRUE,
  arrow_size = 1,
  arrow_position = "tl",
  scalebar = TRUE,
  scalebar_size = 1.5,
  scalebar_position = "tl",
  plot_title = "Mapmixture Figure Format 1",
  plot_title_size = 12,
  axis_title_size = 10,
  axis_text_size = 8
)
```

Run `mapmixture()` with using admixture format 3 and custom parameters:

```{r}
# Load package
library(mapmixture)

# Admixture Format 3
file <- system.file("extdata", "admixture3.csv", package = "mapmixture")
admixture3 <- read.csv(file)

# Coordinates
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)

mapmixture(admixture3, coordinates,
  cluster_cols = c("blue","green"),
  cluster_names = c("Group 1","Group 2"),
  crs = 3035,
  boundary = c(xmin=-15, xmax=16, ymin=40, ymax=62),
  pie_size = 1,
  pie_border = 0.3,
  pie_opacity = 1,
  land_colour = "#d9d9d9",
  sea_colour = "#deebf7",
  expand = TRUE,
  arrow = TRUE,
  arrow_size = 1,
  arrow_position = "tl",
  scalebar = TRUE,
  scalebar_size = 1.5,
  scalebar_position = "tl",
  plot_title = "Mapmixture Figure Format 3",
  plot_title_size = 12,
  axis_title_size = 10,
  axis_text_size = 8
)
```

Run `structure_plot()` to plot a traditional STRUCTURE barplot:

```{r}
# Load package
library(mapmixture)

# Admixture Format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

structure_plot(admixture1, type = "structure")
```

Run `structure_plot()` to plot a facet barplot:

```{r}
# Load package
library(mapmixture)

# Admixture Format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)

structure_plot(admixture1, type = "facet", facet_col = 3)
```

## Shiny web server

https://tomjenkins.shinyapps.io/mapmixture/

## Format

```{r}
# Load package
library(mapmixture)

# Admixture Format 1
file <- system.file("extdata", "admixture1.csv", package = "mapmixture")
admixture1 <- read.csv(file)
head(admixture1)

# Admixture Format 2
file <- system.file("extdata", "admixture2.csv", package = "mapmixture")
admixture2 <- read.csv(file)
admixture2

# Admixture Format 3
file <- system.file("extdata", "admixture3.csv", package = "mapmixture")
admixture3 <- read.csv(file)
admixture3

# Coordinates
file <- system.file("extdata", "coordinates.csv", package = "mapmixture")
coordinates <- read.csv(file)
coordinates
```
